Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-17 MScDiplom_Lemenkova.tex
Some final improvements

\phantomsection	
\addcontentsline{toc}{section}{Front Page}
\begin{titlepage}
	\centering
	\title{Seagrass mapping and monitoring \\ along the coasts of Crete, Greece}
	\author[1]{by Lemenkova Polina}
	%\affil[1]{MSc student at the \textit{University of Twente, Faculty of Geo-Information Science and Earth Observation (ITC)}, The Netherlands}
	\date{\today}
	\maketitle	
\end{titlepage}
\pagebreak

@@

\subsection{Research objective}\index[concepts]{Research!objective}
The current MSc research aims to explore the environmental conditions for the spatial distribution of \textit{P.oceanica} seagrass along the northern coast of Crete Island, based on the remote sensing and \ac{GIS} techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources \index[concepts]{Data}:
	\begin{itemize}
		\item[] i) spectra of \textit{P.oceanica}, carbonate sand, silt and other seafloor types
		\item[] ii) satellite imagery: \ac{Landsat TM}, \ac{Landsat ETM+}
		\item[] iii) aerial photos: Google Earth
		\item[] iv) \textit{in-situ} fieldwork data of underwater videographic measurements
		\item[] v) vector GIS layers.
	\end{itemize}
Although there are a variety of environmental factors that contribute to the spectral
reflectance, the most important ones are water column height and seafloor fraction.
It is because spectra of the seagrass \textit{P.oceanica} vary qualitatively over the depths interval of 0.5–4 m,
and secondly, the content and cover fraction of the seafloor have the most distinctive effect on the
spectral reflectance of the water. 
In the environmental conditions of Crete Island, \textit{P.oceanica} may grow at depths up to six meters; however, the most usual depth is four meters, 
which caused our decision to focus on depths up to four meters, and use these depths for further WASI modelling. 
Therefore, in the current research we focus on these two major factors, and study the response of the
water reflectance towards changing conditions of the water column depth and seafloor bottom cover
fraction (seagrass \textit{P.oceanica} and carbonate sand).\\
The research aims to explore the limitations of the application of the \ac{RTM} and remote sensing techniques 
towards the study of the environmental properties of the \textit{P.oceanica} meadows distribution
over the selected locations of northern Crete. The main research objective is monitoring the seagrass \textit{P.oceanica} in selected areas of Crete and analyzing the environmental conditions for the growth of \textit{P.oceanica}. \\
This research is supported by the \textit{in-situ} measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: spectral modelling by means of the \ac{RTM} \ac{WASI}, underwater videometric measurements made by \ac{Olympus ST} camera, aerial Google Earth and satellite images from different sources, spatial \ac{GIS} and statistical analysis.

@@

\subsubsection{Specific objectives}\index[concepts]{Research!objective!specific}
	\begin{itemize}	
		\item[]1.	To study narrow-band spectral reflectance properties of \textit{P.oceanica} and other seafloor cover types (sand and silt) using \ac{WASI} water colour simulation software 
		\item[]2.	To use methods of the in situ diving observations and underwater videometric measurements by \ac{Olympus ST} camera in order to receive large-scale imagery of the \textit{P.oceanica} mattes 
		\item[]3.	To apply remote sensing data (Google Earth aerial images, \ac{Landsat TM} and \ac{Landsat ETM+} satellite images) for the monitoring of the seagrass meadows distribution 
		\item[]4.	To perform supervised images classification for the thematic mapping of the \textit{P.oceanica} seagrass distribution along the selected locations over the coasts of northern Crete. 
	\end{itemize}
\pagebreak

@@

\subsection{Assumptions}\index[concepts]{Research!assumptions}
The general research assumption, used in this work, in order to make feasible application of the remote sensing methods, includes some statements about the \textit{in-situ} atmospheric conditions, viewing angle, wave backscattering and other optical properties of the environment. They all definitely play a certain role and impact final results of optical measurements, but for practical reasons we have chosen to ignore their contribution in the spectral separability of seafloor cover types. We briefly list below general research assumptions for the optical properties of the environmental variables used in this study. \vspace{1ex}\\
Weather conditions for the measurements are assumed to be ideal: clear, sunny, windless day. Otherwise wind roughens water surface causing sun glitter, and values of spectral reflectance may contain disturbances. \\ Sky radiance might be influenced by multiple reflectance between the sea surface albedo and the atmosphere and in general, sky radiance increases rapidly while viewing zenith angle is near to $90\,^{\circ}$ (i.e. in the evening hours). Therefore, ideally measurements should be done at noon, with as low zenith angle (Fig.\ref{fig:A.2-1})\label{zenith} as possible; otherwise, if the solar zenith angle is too high (e.g. approaching $60-70\,^{\circ}$), received data may contain noise. For \ac{WASI} simulations reflection factor of sky radiance is taken as 0.0201 with simulated ideal conditions: viewing angle ($0\,^{\circ}$ = nadir). The interval of water temperature taken as default lies in the diapason $17-25\,^{\circ}$ as to simulate the conditions of the Mediterranean Sea. However, in real time conditions the sun zenith angle of $45\,^{\circ}$ has been accepted as suitable. 
The accepted model-specific parameters of the \ac{WASI} simulator are shown in the Table \ref{tab:3}.\label{page-10} \vspace{1ex}\\
The anisotropy factor of upwelling radiation or the quality (Q-) factor, showing the directionally dependency of the radiance, is taken as 5. We accepted some values of model-specific optical parameters as default values at \ac{WASI} simulator, which are shown in the Table \ref{tab:3}. Thus, concentration of phytoplankton is accepted at the interval of $0.035-0.089 mg^{-1}$ and concentration of large suspended particles is given to 8. Reference wavelength for \ac{CDOM}, or Gelbstoff, absorption is equal to 440. The backscattering (Fig.\ref{fig:backscat}) is accepted to be $0.00144m^{-1}$. The coefficient of attenuation remains equal to 1.0546, as set up by default at \ac{WASI}.\vspace{1ex}\\
We also assume that concentration of non-chlorophyll particles (absorption at 0) as well as concentration of small suspended particles is equal to zero, so we do not count them in this work. Exponent of \ac{CDOM} (Gelbstoff) absorption is accepted as 0.0140. Finally, the \ac{BRDF} of bottom reflectance (sand), which defines the reflection of light  is at an opaque surface (Fig.\ref{fig:BRDF})\label{page-11}, is assumed to be 0.318 $sr^{-1}$.

\subsection{Research approach}\index[concepts]{Research!approach}
Seagrass consistent monitoring and mapping is necessary and important for the sustainable coastal
development and conservation measures. Earlier, many seagrass meadows have been destroyed by
human activities in the coastal zone, mainly due to the ignorance of their existence, because
information on the seagrass bed exact location was not available \cite{Choo06}\label{Choo06}. Well-time seagrass
observations and mapping enables precise control of its spatial distribution, detection of any changes
in the seagrass landscapes, highlights potential environmental threats in the coastal zone (e.g.
declining of meadows) before they become unmanageable for the coastal management services.
Choosing the right and most effective approach method for the seagrass monitoring is essential.
Remote sensing methods alone, though having evident advantages, are insufficient, because satellite
images of underwater habitats are notoriously difficult to identify and interpret. The best research
method should be based on the integrated approach, well described in various scientific works \cite{Brown02,Montefalcone06,Kirkman96}\label{Brown02} \label{Montefalcone06} \label{Kirkman96}, which includes combination of various
techniques of the seagrass monitoring, i.e. remote sensing imagery classification of aerial and satellite
images, GIS-based spatial analysis and ground \textit{in-situ} surveys.
The current study is based on the application of the remote sensing data, broadband satellite imagery,
aerial images and the results of the underwater videographic measurements towards seagrass
mapping  (Fig.\ref{fig:1.11})\label{page-9}. Image classification is based on the principle of the differentiation between the spectral
signatures of various seafloor cover types  (Fig.\ref{fig:1.10})\label{page-9}. The spectrum of light coming up from the ocean surface in
shallow waters keeps information on the optical properties of the seawater components and benthic
substrate which can be read from their spectral signatures \cite{Werdell03}\label{Werdell03}.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.63]{UML_ClassDiagram.jpg}
	\caption{General methodological approach for the analysis of spectral signatures. \href{http://argouml.tigris.org/}{ArgoUML}, \href{http://www.gimp.org/}{Gimp}. \\ Page: $<$\pageref{page-9}$>$}
	\label{fig:1.10}
\end{figure}

The pre-processing of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution (Fig.\ref{fig:A.50})\label{GG-2}, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).
\pagebreak

@@

%
\subsection{Analysis of spectral signatures}
The distinguishing spectral signatures for various seafloor types (e.g. seagrass species, coral reefs,
various types of sand, mud, other sediments, (Fig.\ref{fig:4.4})\label{page-31} exist in well-defined and narrow (10–20 nm)
wavelength ranges. 
The values of their spectral reflectance are accepted as constant. The results of
spectral measurements enable to analyze, whether \textit{P.oceanica} is spectrally distinct from other sea
floor types with changing environmental conditions, using the differences in their spectral signatures
on the graphs in a \ac{WASI} software. \\The \ac{WASI}, a
software tool for analyzing and simulating the most common types of spectra \cite{Gege05}\label{Gege05}, is highly
suitable for the seagrass spectral analysis.
There are several environmental characteristics, included in \ac{WASI} interface (Fig.\ref{fig:4.7}), which influence the
results of water spectral reflectance, e.g. different bottom depths, concentration of suspended particles
in water column, water temperature, sun angle, concentration of Gelbstoff (\ac{CDOM}), concentration of phytoplankton, aerosol scattering, exponent of backscattering by small particles \cite{Gege04}\label{Gege04}.

@@

at each data set (for example, Fig.\ref{fig:A.11}) with summary outcome of mean, median, Q1 and Q3, standard deviation, min and max values (Tab.\ref{tab:4}, Tab.\ref{tab-16}).\label{page-sumstat} 

@@

As the observations were made in a repeated way, we also analyzed the data pool using autocorrelation as well: Tab.\ref{tab:13} and Tab.\ref{tab:14} for \textit{P.oceanica} and Tab.\ref{tab:28} for carbonate sand. The \textit{autocorrelation}\index[concepts]{Data!statistical analysis!autocorrelation}, a cross-correlation\index[concepts]{Data!statistical analysis!cross-correlation} of an observed value with itself, made on two different days (14th and 15th October), shows the similarity between observations as a function of the time separation between them. 
As can be seen on Fig.\ref{fig:A.21}\label{A-23}, Fig.\ref{fig:A.22}\label{A-24} and Fig.\ref{fig:A.29}, it is a useful tool for finding repeating patterns for a single value of observations, such as the presence of a periodic signal which has been buried under noise. The autocorrelation analysis has been performed using both \ac{SPSS} and \ac{GRETL} software (Fig.\ref{tab:28}\label{page-a-c}). 

@@

To estimate spectral density of a spectral reflectance measurements we used periodogram\index[concepts]{Data!statistical analysis!periodogram} function (Tab.\ref{tab:30})\label{page-period}, which corresponds to the general spectrum of the observations with representation of a variable quantity.  
Thus, the periodogram (Fig.\ref{fig:A.31})\label{page-A31} of the series of observation\index[concepts]{Data!statistical analysis!observations, series of} highlights so-called “dead spots” of low power between the frequencies we wish to exclude (can be seen clearly at the end of the graphs, e.g. Fig.\ref{fig:A.12} ) 
and the frequencies we want to retain (the general profile that lies in the 450-850 nm, Fig.\ref{fig:A.14}, enlarged part of 500-660nm).\\

@@

areas of quartiles\index[concepts]{Data!statistical analysis!quartiles Q1-Q3} (Q1 - Q3, Fig.\ref{fig:A.30}\label{page-A30}, and shaded Q-Q areas on graph, Fig. \ref{fig:4.10}, Fig.\ref{fig:4.14}), which cover most probable data distribution (spectral reflectance) for each spectral band. In x-axis displayed are the areas of 400 – 950 nm in spectra, where the measurements were done; the y-axis was adjusted for the better visualization of the

@@

\paragraph{ANOVA-testing}\index[concepts]{Data!statistical analysis!ANOVA}
We applied \ac{ANOVA} (ANalysis Of Variance) testing for the analysis of the probability of the
spectral reflectance of \textit{P.oceanica} to be more or less spectrally distinct from other seafloor cover
types with changing depth. W are checking the statement Hypothesis $H_o$, is true, which is “the
spectral distinguishability of the seagrass \textit{P.oceanica} from other seafloor types is not changing with
varying \textit{in-situ} conditions, $H_o: \mu1 =\mu2 =\mu3=...= \mu n$, which gives the following outcome.
P(reject $H_o\mid H_o$ is valid)= P(X$>c\mid p$=.05).
where c comes for critical values and p=probability.
The result P=.05 is (Tab.\ref{tab:8}), hence, very small, which makes the statement of Hypothesis $H_o$ highly unlikely
(less than 1 in a 10 chance). The one-way \ac{ANOVA} highlighted a significant difference between data
of spectral reflectances of \textit{P.oceanica} and carbonate sand at different depths, and and proved spectral discernibility of seagrass \textit{P.oceanica} from carbonate sand (Fig.\ref{fig:4.7})\label{firstq}, i.e. true is the opposite statement: Hypothesis $H_a$=“the spectral discernibility of the seagrass \textit{P.oceanica} is distinctly changing, 
and seagrass can be spectrally discriminated from carbonate sand with varying \textit{in-situ} conditions,
 i.e. increasing depth, $H_o: \mu\neq\mu2\neq\mu3\neq ... \neq\mu n$.
As a result of the statistical testing, we came to the following conclusion.
The Hypothesis $H_a$ is true, which claims that the spectral discernibility of the seagrass \textit{P.oceanica} is
distinctly changing and can be discriminated from from other seafloor cover types (carbonate sand) with varying \textit{in-situ} conditions, 
i.e. increasing depth, $H_o: \mu1\neq\mu2\neq\mu3\neq...\neq\mu n$, which positively answers first research question.
Statistical results are illustrated with Error bars made using Excel's vertical box and Whisker Charts (Box Plots).
The graphs were plotted on the basis of the following statistical data calculated from the sampling
measurements data: Table 2.

\begin{table}
\rowcolors{1}{Khaki}{LightYellow}
\caption{Statistical analysis of the measurements of spectral reflectance of \textit{P.oceanica} (fragment). Page: $<$\pageref{page-40}$>$}
\centering
  \begin{tabular}{| c | c | c | c | c |}
    \hline
   	 \textbf{Statistical values} & \textbf{318.19} & \textbf{418.3448} & \textbf{518.8361} & \textbf{619.3662} \\ \hline \hline
	    Mean & 0.034521 & 0.049235 & 0.076088 & 0.085287 \\ \hline
	    St Dev & 0.019849 & 0.030338 & 0.042587 & 0.046569 \\ \hline
	   Median & 0.033035 & 0.046197 & 0.07704 & 0.080277 \\ \hline
	   Q1 & 0.019293 & 0.023706 & 0.045606& 0.049468 \\ \hline
	   Q3 & 0.042477 & 0.057934 & 0.092289 & 0.114191 \\ \hline
	   Minimum & 0.007898 & 0.010087 & 0.021373 & 0.022691 \\ \hline
	   Maximum & 0.076835 & 0.110906 & 0.18938 & 0.216231 \\ \hline
	   25th Pct & 0.019293 & 0.023706 & 0.045606 & 0.049468 \\ \hline
	   50th Pct & 0.013743 & 0.022491 & 0.031435 & 0.030808 \\ \hline
	   75th Pct & 0.009441 & -0.01174 & -0.01525 & -0.03391 \\ \hline
	   Min & 0.011395 & 0.013618 & 0.024232 & 0.026777 \\ \hline
	   Max & 0.034358 & 0.052972 & 0.097091 & 0.1 \\ \hline
  \end{tabular}
   \label{tab:4}
\end{table}

where the abbreviations stand for the following values:
Mean: Average of the data to be plotted, AVERAGE\index[concepts]{Data!statistical analysis!average} {data}
St Dev: Standard deviation\index[concepts]{Data!statistical analysis!standard deviation}, \ac{STDEV} {data}
Median: Median of the data, MEDIAN{data}
Calculating interquartile ranges: Q1=First quartile, PERCENTILE({data}*0.25) and Q3=Third
quartile, PERCENTILE({data}*0.75)
Minimum: Minimum value,MIN{data}
Maximum: Maximum value, MAX{data}
25th Pct: Plotting value of first quartile = Q1
50th Pct: Plotting value of median = Median\index[concepts]{Data!statistical analysis!median}-Q1
75th Pct: Plotting value of third quartile = Median-Q3
Min: Lower error bar length=Q1 - Minimum
Max: Upper error bar length=Maximum - Q3
The statistical analysis is displayed on the Graph 1.1 which compares the spread in sets of
measurement data of spectral reflectance of under various environmental conditions (depth).

>>>>

\phantomsection	
\addcontentsline{toc}{section}{Front Page}
\begin{titlepage}
	\centering
	\title{Seagrass mapping and monitoring \\ along the coasts of Crete, Greece}
	\author[1]{by Lemenkova Polina}
	%\affil[1]{MSc student at the \textit{University of Twente, Faculty of Geo-Information Science and Earth Observation (ITC)}, The Netherlands}
	\date{March 2011}
	\maketitle	
\end{titlepage}
\pagebreak

@@

\subsection{Research objective}\index[concepts]{Research!objective}
The current MSc research aims to explore the environmental conditions for the spatial distribution of \textit{P.oceanica} seagrass along the northern coast of Crete Island, based on the remote sensing and \ac{GIS} techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources \index[concepts]{Data}:
	\begin{itemize}
		\item[] i) spectra of \textit{P.oceanica}, carbonate sand, silt and other seafloor types
		\item[] ii) satellite imagery: \ac{Landsat TM}, \ac{Landsat ETM+}
		\item[] iii) aerial photos: Google Earth
		\item[] iv) \textit{in-situ} fieldwork data of underwater videographic measurements
		\item[] v) vector GIS layers.
	\end{itemize}
Although there are a variety of environmental factors that contribute to the spectral
reflectance, the most important ones are water column height and seafloor fraction.
It is because spectra of the seagrass \textit{P.oceanica} vary qualitatively over the depths interval of 0.5–4 m,
and secondly, the content and cover fraction of the seafloor have the most distinctive effect on the
spectral reflectance of the water. 
In the environmental conditions of Crete Island, \textit{P.oceanica} may grow at depths up to six meters; however, the most usual depth is four meters, 
which caused our decision to focus on depths up to four meters, and use these depths for further WASI modelling. 
Therefore, in the current research we focus on these two major factors, and study the response of the
water reflectance towards changing conditions of the water column depth and seafloor bottom cover
fraction (seagrass \textit{P.oceanica} and carbonate sand).\\
The research aims to explore the limitations of the application of the \ac{RTM} and remote sensing techniques 
towards the study of the environmental properties of the \textit{P.oceanica} meadows distribution
over the selected locations of northern Crete. The main research objective is monitoring the seagrass \textit{P.oceanica} in selected areas of Crete and analyzing the environmental conditions for the growth of \textit{P.oceanica}. \\
This research is supported by the \textit{in-situ} measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: spectral modelling by means of the \ac{RTM} \ac{WASI}, underwater videometric measurements made by \ac{Olympus ST} camera, aerial Google Earth and satellite images from different sources, spatial \ac{GIS} and statistical analysis.
\pagebreak

@@

\subsubsection{Specific objectives}\index[concepts]{Research!objective!specific}
	\begin{itemize}	
		\item[]1.	To study narrow-band spectral reflectance properties of \textit{P.oceanica} and other seafloor cover types (sand and silt) using \ac{WASI} water colour simulation software 
		\item[]2.	To use methods of the in situ diving observations and underwater videometric measurements by \ac{Olympus ST} camera in order to receive large-scale imagery of the \textit{P.oceanica} mattes 
		\item[]3.	To apply remote sensing data (Google Earth aerial images, \ac{Landsat TM} and \ac{Landsat ETM+} satellite images) for the monitoring of the seagrass meadows distribution 
		\item[]4.	To perform supervised images classification for the thematic mapping of the \textit{P.oceanica} seagrass distribution along the selected locations over the coasts of northern Crete. 
	\end{itemize}

@@

\subsection{Assumptions}\index[concepts]{Research!assumptions}
The general research assumption, used in this work, in order to make feasible application of the remote sensing methods, includes some statements about the \textit{in-situ} atmospheric conditions, viewing angle, wave backscattering and other optical properties of the environment. They all definitely play a certain role and impact final results of optical measurements, but for practical reasons we have chosen to ignore their contribution in the spectral separability of seafloor cover types. We briefly list below general research assumptions for the optical properties of the environmental variables used in this study. \vspace{1ex}\\
Weather conditions for the measurements are assumed to be ideal: clear, sunny, windless day. Otherwise wind roughens water surface causing sun glitter, and values of spectral reflectance may contain disturbances. \\ Sky radiance might be influenced by multiple reflectance between the sea surface albedo and the atmosphere and in general, sky radiance increases rapidly while viewing zenith angle is near to $90\,^{\circ}$ (i.e. in the evening hours). Therefore, ideally measurements should be done at noon, with as low zenith angle (Fig.\ref{fig:A.2-1})\label{zenith} as possible; otherwise, if the solar zenith angle is too high (e.g. approaching $60-70\,^{\circ}$), received data may contain noise. For \ac{WASI} simulations reflection factor of sky radiance is taken as 0.0201 with simulated ideal conditions: viewing angle ($0\,^{\circ}$ = nadir). The interval of water temperature taken as default lies in the diapason $17-25\,^{\circ}$ as to simulate the conditions of the Mediterranean Sea. However, in real time conditions the sun zenith angle of $45\,^{\circ}$ has been accepted as suitable. 
The accepted model-specific parameters of the \ac{WASI} simulator are shown in the Table \ref{tab:3}.\label{page-10} \vspace{1ex}\\
The anisotropy factor of upwelling radiation or the quality (Q-) factor, showing the directionally dependency of the radiance, is taken as 5. We accepted some values of model-specific optical parameters as default values at \ac{WASI} simulator, which are shown in the Table \ref{tab:3}. Thus, concentration of phytoplankton is accepted at the interval of $0.035-0.089 mg^{-1}$ and concentration of large suspended particles is given to 8. Reference wavelength for \ac{CDOM}, or Gelbstoff, absorption is equal to 440. The backscattering (Fig.\ref{fig:backscat}) is accepted to be $0.00144m^{-1}$. The coefficient of attenuation remains equal to 1.0546, as set up by default at \ac{WASI}.\vspace{1ex}\\
We also assume that concentration of non-chlorophyll particles (absorption at $\lambda_0$) as well as concentration of small suspended particles is equal to zero, so we do not count them in this work. Exponent of \ac{CDOM} (Gelbstoff) absorption is accepted as 0.0140. Finally, the \ac{BRDF} of bottom reflectance (sand), which defines reflection of light  is at an opaque surface (Fig.\ref{fig:BRDF})\label{page-11}, is assumed to be 0.318 $sr^{-1}$.

\subsection{Research approach}\index[concepts]{Research!approach}
Seagrass consistent monitoring and mapping is necessary and important for the sustainable coastal
development and conservation measures. Earlier, many seagrass meadows have been destroyed by
human activities in the coastal zone, mainly due to the ignorance of their existence, because
information on the seagrass bed exact location was not available \cite{Choo06}\label{Choo06}. Well-time seagrass
observations and mapping enables precise control of its spatial distribution, detection of any changes
in the seagrass landscapes, highlights potential environmental threats in the coastal zone (e.g.
declining of meadows) before they become unmanageable for the coastal management services.
Choosing the right and most effective approach method for the seagrass monitoring is essential.
Remote sensing methods alone, though having evident advantages, are insufficient, because satellite
images of underwater habitats are notoriously difficult to identify and interpret. The best research
method should be based on the integrated approach, well described in various scientific works \cite{Brown02,Montefalcone06,Kirkman96}\label{Brown02} \label{Montefalcone06} \label{Kirkman96}, which includes combination of various
techniques of the seagrass monitoring, i.e. remote sensing imagery classification of aerial and satellite
images, GIS-based spatial analysis and ground \textit{in-situ} surveys.
The current study is based on the application of the remote sensing data, broadband satellite imagery,
aerial images and the results of the underwater videographic measurements towards seagrass
mapping  (Fig.\ref{fig:1.11})\label{page-9}. Image classification is based on the principle of the differentiation between the spectral
signatures of various seafloor cover types  (Fig.\ref{fig:1.10})\label{page-9}. The spectrum of light coming up from the ocean surface in
shallow waters keeps information on the optical properties of the seawater components and benthic
substrate which can be read from their spectral signatures \cite{Werdell03}\label{Werdell03}. 
The pre-processing of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution (Fig.\ref{fig:A.50})\label{GG-2}, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.63]{UML_ClassDiagram.jpg}
	\caption{General methodological approach for the analysis of spectral signatures. \href{http://argouml.tigris.org/}{ArgoUML}, \href{http://www.gimp.org/}{Gimp}. \\ Page: $<$\pageref{page-9}$>$}
	\label{fig:1.10}
\end{figure}
\pagebreak

@@

%
\subsection{Analysis of spectral signatures}
The distinguishing spectral signatures for various seafloor types (e.g. seagrass species, coral reefs,
various types of sand, mud, other sediments, (Fig.\ref{fig:4.4})\label{page-31} exist in well-defined and narrow ($10 - 20$ nm)
wavelength ranges. 
The values of their spectral reflectance are accepted as constant. The results of
spectral measurements enable to analyze, whether \textit{P.oceanica} is spectrally distinct from other sea
floor types with changing environmental conditions, using the differences in their spectral signatures
on the graphs in a \ac{WASI} software. \\The \ac{WASI}, a
software tool for analyzing and simulating the most common types of spectra \cite{Gege05}\label{Gege05}, is highly
suitable for the seagrass spectral analysis.
There are several environmental characteristics, included in \ac{WASI} interface (Fig.\ref{fig:4.7}), which influence the
results of water spectral reflectance, e.g. different bottom depths, concentration of suspended particles
in water column, water temperature, sun angle, concentration of Gelbstoff (\ac{CDOM}), concentration of phytoplankton, aerosol scattering, exponent of backscattering by small particles \cite{Gege04}\label{Gege04}.

@@

at each data set (for example, Fig.\ref{fig:A.11}) with summary outcome of mean, median, Q1 and Q3, standard deviation, min and max values (Tab.\ref{tab:4}, Tab.\ref{tab:A.8}).\label{page-sumstat} 

@@

As the observations were made in a repeated way, we also analyzed the data pool using autocorrelation as well: Tab.\ref{tab:13} and Tab.\ref{tab:14} for \textit{P.oceanica} and Tab.\ref{tab:28} for carbonate sand. The \textit{autocorrelation}\index[concepts]{Data!statistical analysis!autocorrelation}, a cross-correlation\index[concepts]{Data!statistical analysis!cross-correlation} of an observed value with itself, made on two different days (14th and 15th October), shows the similarity between observations as a function of the time separation between them, 
as can be seen on Fig.\ref{fig:A.21}\label{A-23}, Fig.\ref{fig:A.22}\label{A-24} and Fig.\ref{fig:A.29}, it is a useful tool for finding repeating patterns for a single value of observations, such as the presence of a periodic signal which has been buried under noise. The autocorrelation analysis has been performed using both \ac{SPSS} and \ac{GRETL} software (Fig.\ref{tab:28}\label{page-a-c}). 

@@

To estimate spectral density of optical measurements we used periodogram\index[concepts]{Data!statistical analysis!periodogram} function (Tab.\ref{tab:30})\label{page-period}, which corresponds to the general spectrum of the observations with representation of a variable quantity.  
Thus, the periodogram (Fig.\ref{fig:A.31})\label{page-A31} of the series of observation\index[concepts]{Data!statistical analysis!observations, series of} highlights so-called “dead spots” of low power between the frequencies we wish to exclude (can be seen clearly at the end of the graphs, e.g. Fig.\ref{fig:A.12}) and the frequencies we want to retain (the general profile that lies in the 450-850 nm, Fig.\ref{fig:A.14}, enlarged part of 500-660nm).\\

@@

areas of quartiles\index[concepts]{Data!statistical analysis!quartiles Q1-Q3} (Q1 - Q3, Fig.\ref{fig:A.30}\label{page-A30}, and shaded Q-Q areas on graph, Fig. \ref{fig:4.10}, Fig.\ref{fig:4.14}), which cover most probable data distribution (spectral reflectance) for each spectral band. In x-axis displayed are the areas of 400-950 nm in spectra, where the measurements were done; the y-axis was adjusted for the better visualization of the

@@

\paragraph{ANOVA-testing}\index[concepts]{Data!statistical analysis!ANOVA}
We applied \ac{ANOVA} (ANalysis Of Variance) testing for the analysis of the probability of the
spectral reflectance of \textit{P.oceanica} to be more or less spectrally distinct from other seafloor cover
types with changing depth. W are checking the statement Hypothesis $H_o$, is true, which is “the
spectral distinguishability of the seagrass \textit{P.oceanica} from other seafloor types is not changing with
varying \textit{in-situ} conditions, $H_o: \mu1 =\mu2 =\mu3=...= \mu n$, which gives the following outcome.
P(reject $H_o\mid H_o$ is valid)= P(X$>c\mid p$=.05).
where c comes for critical values and p=probability.
The result P=.05 is (Tab.\ref{tab:8}), hence, very small, which makes the statement of Hypothesis $H_o$ highly unlikely
(less than 1 in a 10 chance). The one-way \ac{ANOVA} highlighted a significant difference between data
of spectral reflectances of \textit{P.oceanica} and carbonate sand at different depths, and and proved spectral discernibility of seagrass \textit{P.oceanica} from carbonate sand (Fig.\ref{fig:4.7})\label{firstq}, i.e. true is the opposite statement: Hypothesis $H_a$=“the spectral discernibility of the seagrass \textit{P.oceanica} is distinctly changing, 
and seagrass can be spectrally discriminated from carbonate sand with varying \textit{in-situ} conditions,
 i.e. increasing depth, $H_o: \mu\neq\mu2\neq\mu3\neq ... \neq\mu n$.
As a result of the statistical testing, we came to the following conclusion.
The Hypothesis $H_a$ is true, which claims that the spectral discernibility of the seagrass \textit{P.oceanica} is
distinctly changing and can be discriminated from from other seafloor cover types (carbonate sand) with varying \textit{in-situ} conditions, 
i.e. increasing depth, $H_o: \mu1\neq\mu2\neq\mu3\neq...\neq\mu n$, which positively answers first research question.
Statistical results are illustrated with Error bars made using Excel's vertical box and Whisker Charts (Box Plots).
The graphs were plotted on the basis of the following statistical data calculated from the sampling
measurements data: Tab.\ref{tab:A.8}, Tab.\ref{tab:A.29}, Tab\ref{tab:4.1}.\label{last}

\begin{table}[H]
\rowcolors{1}{Khaki}{LightYellow}
\caption{Statistical analysis of the measurements of spectral reflectance of \textit{P.oceanica} (fragment). Page: $<$\pageref{page-40}$>$}
\centering
  \begin{tabular}{| c | c | c | c | c |}
    \hline
   	 \textbf{Statistical values} & \textbf{318.19} & \textbf{418.3448} & \textbf{518.8361} & \textbf{619.3662} \\ \hline \hline
	    Mean & 0.034521 & 0.049235 & 0.076088 & 0.085287 \\ \hline
	    St Dev & 0.019849 & 0.030338 & 0.042587 & 0.046569 \\ \hline
	   Median & 0.033035 & 0.046197 & 0.07704 & 0.080277 \\ \hline
	   Q1 & 0.019293 & 0.023706 & 0.045606& 0.049468 \\ \hline
	   Q3 & 0.042477 & 0.057934 & 0.092289 & 0.114191 \\ \hline
	   Minimum & 0.007898 & 0.010087 & 0.021373 & 0.022691 \\ \hline
	   Maximum & 0.076835 & 0.110906 & 0.18938 & 0.216231 \\ \hline
	   25th Pct & 0.019293 & 0.023706 & 0.045606 & 0.049468 \\ \hline
	   50th Pct & 0.013743 & 0.022491 & 0.031435 & 0.030808 \\ \hline
	   75th Pct & 0.009441 & -0.01174 & -0.01525 & -0.03391 \\ \hline
	   Min & 0.011395 & 0.013618 & 0.024232 & 0.026777 \\ \hline
	   Max & 0.034358 & 0.052972 & 0.097091 & 0.1 \\ \hline
  \end{tabular}
   \label{tab:4}
\end{table}

The abbreviations stand for the following values:
Mean: Average of the data to be plotted, AVERAGE\index[concepts]{Data!statistical analysis!average} {data}
St Dev: Standard deviation\index[concepts]{Data!statistical analysis!standard deviation}, \ac{STDEV} {data}
Median: Median of the data, MEDIAN{data}
Calculating interquartile ranges: Q1=First quartile, PERCENTILE({data}*0.25) and Q3=Third
quartile, PERCENTILE({data}*0.75)
Minimum: Minimum value,MIN{data}
Maximum: Maximum value, MAX{data}
25th Pct: Plotting value of first quartile = Q1
50th Pct: Plotting value of median = Median\index[concepts]{Data!statistical analysis!median}-Q1
75th Pct: Plotting value of third quartile = Median-Q3
Min: Lower error bar length=Q1 - Minimum
Max: Upper error bar length=Maximum - Q3
The statistical analysis is displayed on the Graph 1.1 which compares the spread in sets of
measurement data of spectral reflectance of under various environmental conditions (depth).
